---
title: "Notes on R for Data Science Chapter 5: Data transformation, Part 2"
output: html_notebook
---

*Ideas and exercises come from https://r4ds.had.co.nz/transform.html*

*Additional notes by TCS*

# Setup
First, we load the `tidyverse` package and a dataset. This data frame contains all 336,776 flights that departed from New York City in 2013.

```{r setup}
require(nycflights13)
require(tidyverse)
require(hms)
flights
```

# Review: Basic dplyr functions

There are 5 key functions ("verbs"), plus a helper function, that do most data manipulation tasks in dplyr:  

* `filter`: pick observations by values
* `arrange`: reorder rows  
* `select`: pick variables by name  
* `mutate`: create new variables from existing ones, using functions  
* `summarise`: collapse values into single ones  
* `group_by`: change scope of a verb from the whole dataset to individual groups 

## `filter()` gives you a subset of rows based on values

## NA handling in `filter`

`filter()` includes ONLY rows where the condition is TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly:  

```{r filter NA examples, echo = TRUE}
df <- tibble(x = c(1, NA, 3))
(biggerthan1 <- filter(df, x > 1))
(bigorNA <- filter(df, is.na(x) | x > 1))
```
# `arrange()` is for sorting rows

* The arguments for `arrange()` are a dataframe and column name(s).   
* The `desc` option reverses the order.  
* Missing values (NAs) are always at the end regardless of `desc()` 
 
# `select()` gives you a subset of columns by name

You can name each column, or specify a range using a colon.  
As with other R selections, you can omit certain columns using the minus sign.  
You can add multiple arguments to include more columns in the selection.  

## Partial names  

`select()` does not have to use exact column matches.  

* `starts_with("foo")`  
* `ends_with("bar")`  
* `contains("foobar")`  
* `matches(some_regex)`  
* `num_range("x", 1:3)` matches x1, x2 and x3  

## Variants on `select()`

* `rename()` is considered a variant of select() where you take a column, change its name, and keep all other columns as well.
* `everything()` is a helper for select() that lets you move one or a few columns to the beginning (left) of the table, while retaining all other columns.  

# `mutate()` is for creating new variables from old  

* The new columns are added at the end (right) of the data frame.  
* You can add multiple columns in one call, and the later columns can use variables defined earlier in the call  
* `transmute()` is a variant where you keep ONLY the newly defined columns  

```{r mutate examples, echo = TRUE}
(flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
))
(flights_less_sml <- mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
))
(flights_even_more <- mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
))
(flights_new <- transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
))
```

## Many vectorized functions work with `mutate()`  

Some examples:  
* Arithmetic, logs (vectors recycle to match length)  
* Modulus arithmetic:  %/% (integer division) and %%   (remainder), where x == y * (x %/% y) + (x %% y)  
* Surrounding values: lead(), lag()  
* Cumulative values such as cumsum() and cumprod()  
* With the Rcpproll package, rolling sums, etc.  
* Ranking such as min_rank()  

## Exercises for `mutate()`  

1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.  

2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?  

*The times are in HHMM format, so straight subtraction does not work. We can convert to time objects, or use modulus arithmetic to convert to minutes and back to time.  

3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?  

*We expect dep_delay = sched_dep_time - dep_time. However, for > 60 min difference, we have the same time issue, and solution, as in exercise 2.*

4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().  

* *`rank()` goes smallest to largest, gives decimals (averages) for ties, 1, 2.5, 2.5, 4,... and puts NAs last by default, but can change behavior using `na.last`.  `rank(desc())` goes largest to smallest.*  
* *`rank()` has different methods of handling ties, producing all-integer output. First; last; random; max (max rank of all ties so there might not be a #1); min (min rank of all tied elements "as used in sports," so you can "tie for 1st" -- (1,1,3,4...)) -- this is min_rank(). `first`, `last`, `random` produce all-unique ranks.*  
* *If you preserve or average ties, you could end up with more or fewer than 10 in the result. Therefore to get exactly 10 we will break ties randomly.* 
* *If you want to use another column to break ties, then I don't see a way within `rank()` functions. You could use `arrange()` and then row numbers.*  

5. What does 1:3 + 1:10 return? Why?  

*The 1:3 vector is recycled, so you get 4+1, 5+2, etc.*  

6. What trigonometric functions does R provide?  

*The usual functions, plus a few others. Angles are in radians. `cospi(x)` and kin give the functions of pi times x, only for x = multiples of 0.5*

```{r mutate exercises, echo =TRUE}
# The time is in HMM or HHMM format. Therefore to get the hours we divide by 100, and for minutes we take the remainder. (Did not deal with the times = 2400!)

(better_time <- mutate(flights, dep_min_since_midnight = (dep_time %/% 100) * 60 + dep_time %% 100, sched_dep_min_since_midnight = (sched_dep_time %/% 100) * 60 + sched_dep_time %% 100))

(compare <- select(flights, air_time, arr_time, dep_time) %>% mutate(spent_time = arr_time-dep_time))

# the times are in HHMM format, so straight subtraction does not work

(realtimes <- mutate(flights, 
                    arr_time_hms = hms(NULL, arr_time %% 100, arr_time %/% 100),
                    dep_time_hms = hms(NULL, dep_time %% 100, dep_time %/% 100),
                    spent_time_hms = difftime(arr_time_hms, dep_time_hms, units = "mins")))
  
# mutate to add the ranking function (alternatively you could probably arrange and then take by row number)

# if you preserve or average ties, you could end up with more or fewer than 10 in the result. Therefore to get exactly 10 we will break ties randomly.

(arr_ranked <- mutate(flights, arr_delay_rank = rank(desc(arr_delay), ties.method = "random")))
(top_10_delay <- arr_ranked %>%  filter(arr_delay_rank <= 10))
```

# `summarise()` collapses groups into single values

