---
title: "Notes on R for Data Science Chapter 5: Data transformation, Part 2"
output: html_notebook
---

*Ideas and exercises come from https://r4ds.had.co.nz/transform.html*

*Additional notes by TCS*

# Setup
First, we load the `tidyverse` package and a dataset. This data frame contains all 336,776 flights that departed from New York City in 2013.

```{r setup}
require(nycflights13)
require(tidyverse)
flights
```

# Review: Basic dplyr functions

There are 5 key functions ("verbs"), plus a helper function, that do most data manipulation tasks in dplyr:  

* `filter`: pick observations by values
* `arrange`: reorder rows  
* `select`: pick variables by name  
* `mutate`: create new variables from existing ones, using functions  
* `summarise`: collapse values into single ones  
* `group_by`: change scope of a verb from the whole dataset to individual groups 

## `filter()` gives you a subset of rows based on values

## NA handling in `filter`

`filter()` includes ONLY rows where the condition is TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly:  

```{r filter NA examples, echo = TRUE}
df <- tibble(x = c(1, NA, 3))
(biggerthan1 <- filter(df, x > 1))
(bigorNA <- filter(df, is.na(x) | x > 1))
```
# `arrange()` is for sorting rows

* The arguments for `arrange()` are a dataframe and column name(s).   
* The `desc` option reverses the order.  
* Missing values (NAs) are always at the end regardless of `desc()` 
 
# `select()` gives you a subset of columns by name

You can name each column, or specify a range using a colon.  
As with other R selections, you can omit certain columns using the minus sign.  
You can add multiple arguments to include more columns in the selection.  

## Partial names  

`select()` does not have to use exact column matches.  

* `starts_with("foo")`  
* `ends_with("bar")`  
* `contains("foobar")`  
* `matches(some_regex)`  
* `num_range("x", 1:3)` matches x1, x2 and x3  

## Variants on `select()`

* `rename()` is considered a variant of select() where you take a column, change its name, and keep all other columns as well.
* `everything()` is a helper for select() that lets you move one or a few columns to the beginning (left) of the table, while retaining all other columns.  

# `mutate()` is for creating new variables from old  

* The new columns are added at the end (right) of the data frame.  
* You can add multiple columns in one call, and the later columns can use variables defined earlier in the call  
* `transmute()` is a variant where you keep ONLY the newly defined columns  

```{r mutate examples, echo = TRUE}
(flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
))
(flights_less_sml <- mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
))
(flights_even_more <- mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
))
(flights_new <- transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
))
```

## Many vectorized functions work with `mutate()`  

Some examples:  
* Arithmetic, logs (vectors recycle to match length)  
* Modulus arithmetic:  %/% (integer division) and %%   (remainder), where x == y * (x %/% y) + (x %% y)  
* Surrounding values: lead(), lag()  
* Cumulative values such as cumsum() and cumprod()  
* With the Rcpproll package, rolling sums, etc.  
* Ranking such as min_rank()  

## Exercises for `mutate()`  

1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.  

2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?  

3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?  

4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().  

5. What does 1:3 + 1:10 return? Why?  

6. What trigonometric functions does R provide?  

```{r mutate exercises, echo =TRUE}

```